Cognate beginnings to bilingual lexical acquisition

### 2023-10-03 manuscript/manuscript.tex
Ignore .tiff, .eps, and .svg

supported by the European Union's Horizon 2023 research and innovation
program under Marie Skłodowska--Curie Grant (765556) and a postdoctoral
fellowship from the Foundation for Science and Technology of Portugal
(UIDB/00214/2020). IC was supported by the Investigo program funded by
the European Union's NextGenerationEU (NGEU) recovery plan. NSG was
supported by an ICREA Academia award from the Catalan Institution for
Research and Advanced Studies (ICREA). We are grateful to Chiara
Santolin, Ege E. Özer, and the rest of the Speech Acquisition and
Perception research group, and to Alicia Franco-Martínez and Cristina
Rodríguez-Prada, for their helpful feedback. We thank Xavier Mayoral,
Silvia Blanch, and Cristina Cuadrado for their technical support, and
Cristina Dominguez and Katia Pistrin for their efforts in recruiting
infants. We also thank all families and infants who participated in the
experiments.}}

@@

08005\\\texttt{\href{mailto:avila.varela.daniela@gmail.com}{avila.varela.daniela@gmail.com}}\And

@@

questionnaire (BVQ), to collect vocabulary data from 366 Catalan-Spanish

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, breakable, frame hidden, boxrule=0pt, interior hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi

@@

Floccia et al. (\protect\hyperlink{ref-floccia2018introduction}{2018})
published an impressive study comparing vocabularies of children
learning several language pairs differing in their percentage of
cognates. The authors collected vocabulary data on word comprehension
and production from 372 24-month-old bilingual toddlers living in the
United Kingdom who were learning English and an additional language. The
additional language was one of 13 typologically diverse languages:
Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi or Urdu,
Italian, Mandarin Chinese, Polish, Portuguese, Spanish, and Welsh. The
authors calculated the average lexical overlap between the words in each
of these additional languages and their translation equivalents in
English. Lexical overlap was calculated in terms of phonological
similarity (described below) and it was taken as a proxy of the degree
of cognateness between each pair of languages. Floccia and co-workers
reported an increase in vocabulary size in the additional language
(i.e., not English) associated with an increase in the average
phonological similarity between the translation equivalents of each
language pair. For example, English-Dutch bilinguals (languages with a
high phonological overlap), were able to produce more Dutch words than
English-Mandarin bilinguals (languages with a low phonological overlap)
were able to produce in Mandarin. Blom et al.
(\protect\hyperlink{ref-blom2020crosslanguage}{2020}), Bosma et al.
(\protect\hyperlink{ref-bosma2019longitudinal}{2019}), and Gampe et al.
(\protect\hyperlink{ref-gampe2021does}{2021}) reported similar results,
providing converging evidence of a facilitatory effect of a lower
language distance (i.e., higher degre of cognateness) on vocabulary
size.

A second set of studies suggested that cognates are overrepresented in
bilinguals' early lexicon. Bosch \& Ramon-Casas
(\protect\hyperlink{ref-bosch2014first}{2014}) collected parental
reports of expressive vocabulary from 48 Catalan-Spanish bilinguals aged
18 months and found that cognates represented a larger proportion of
vocabulary than non-cognates. Schelletter
(\protect\hyperlink{ref-schelletter2002effect}{2002}) provided
converging evidence from a longitudinal single-case study, in which an
English-German bilingual child produced cognates earlier than
non-cognates, on average. But the high proportion of cognates in the
vocabulary of the participants in these two studies may not necessarily
evidence of a facilitation effect of cognateness, but rather of simply
the high proportion of cognates present in the pair of languages being
learned. For instance, if two given languages share a high proportion of
cognates like 70\%, the vocabulary contents of children learning both
languages should, in principle, approximate such proportion of cognates,
even in the absence of a cognateness facilitation effect. More recently,
Mitchell et al. (\protect\hyperlink{ref-mitchell2022cognates}{2022})
addressed this issue in a longitudinal study. The authors collected
expressive vocabulary data of 47 16- to 30-month-old French-English
bilinguals living in Canada, in both languages. They created two lists
of translation equivalents: one made of 131 cognates, and one made of
406 non-cognates. The proportion of words that children were reportedly
able to produce was higher in the cognate lists than in the non-cognate
list across ages. Critically, this difference persisted after both lists
were matched in size, controlling their semantic category (i.e.,
furniture, animals, food were similarity represented in both lists) and
age-of-acquisition norms (an index of word difficulty). Taken together,
the results of these two lines of research have been considered as
supporting to the hypothesis that phonological similarity (as reflected
in cognateness) plays a facilitation role in bilingual word acquisition.

Parallel activation of bilinguals' lexicons has been proposed as the
underlying mechanism for such facilitatory effect (e.g.,
\protect\hyperlink{ref-floccia2018introduction}{Floccia et al., 2018};
\protect\hyperlink{ref-mitchell2022cognates}{Mitchell et al., 2022}).
The parallel activation hypothesis stems from the language non-selective
account of lexical access, which suggests that bilinguals activate both
languages simultaneously during language processing, even in fully
monolingual contexts. Evidence with adult bilinguals supporting the
language-non selective account of lexical access has been reported for
language comprehension and production, across the auditory and visual
(reading and signing) modalities
(\protect\hyperlink{ref-gimeno-martinez2021crosslinguistic}{Gimeno-Martínez
et al., 2021}; \protect\hyperlink{ref-hoshino2008cognate}{Hoshino \&
Kroll, 2008}; \protect\hyperlink{ref-morford2011when}{Morford et al.,
2011}; \protect\hyperlink{ref-shook2012bimodal}{Shook \& Marian, 2012};
\protect\hyperlink{ref-spivey1999cross}{Spivey \& Marian, 1999}; see
\protect\hyperlink{ref-kroll2017bilingual}{Kroll \& Ma, 2017} for
review). One of the clearest pieces of evidence of parallel activation
was provided by Costa et al.
(\protect\hyperlink{ref-costa2000cognate}{2000}). In this study, Spanish
monolinguals and Catalan-Spanish bilingual adults were asked to name
pictures of common objects in Spanish. In half of the trials, the object
labels were cognates in Spanish and Catalan (\emph{árbol}-\emph{arbre},
translations of \emph{tree}), whereas in the other half of the trial
labels were non-cognates (\emph{mesa}-\emph{taula}, translations of
\emph{table}). Obviously, such distinction was only relevant for
bilinguals. Bilinguals named cognate pictures faster than non-cognate
pictures, even after adjusting for the lexical frequency of the items.
In contrast, Spanish monolinguals, who were unfamiliar with the Catalan
translations of the Spanish words they uttered, showed equivalent naming
times for the two types of stimuli. The authors interpreted the
difference between cognates and non-cognates in bilinguals as reflecting
the additional phonological activation that cognate words would receive
from their translation equivalents (due to language non-selective
activation of bilinguals' lexicons). These results showed that
bilinguals' Catalan phonology was activated during the production of
Spanish words, facilitating the naming of cognate pictures. More
recntly, evidence of parallel activation has been reported in bilingual
toddlers and children too
(\protect\hyperlink{ref-bosma2020cognate}{Bosma \& Nota, 2020};
\protect\hyperlink{ref-floccia2020translation}{Floccia et al., 2020};
\protect\hyperlink{ref-poarch2012crosslanguage}{Poarch \& Hell, 2012};
\protect\hyperlink{ref-poulin-dubois2013lexical}{Poulin-Dubois et al.,
2013}; \protect\hyperlink{ref-vonholzen2019impact}{Von Holzen et al.,
2019}; \protect\hyperlink{ref-vonholzen2012language}{Von Holzen \& Mani,
2012}). Although there is a consensus on the role of parallel activation
in bilinguals' lexical processing and acquisition, previous studies do
not address its influence on the learning trajectories of words. Results
are aggregated across words and provide no information about the
specific dynamics of how parallel activation influences word learning.
This is the goal of the present research.

We propose an account in which a learning instance for a word may also
represent a learning instance for its translation equivalent, to the
extent that such translation equivalent is co-activated. We use the term
\emph{learning instance} in the fashion of accumulator models of
language acquisition; as an exposure to a word-form that constitutes an
opportunity for the child to accumulate information about the word. We
do not assume if a learning instance is a discrete or a continuous unit
of accumulation of information. We consider that a learning instance of
a word is an exposure to its (phonological) form if the resulting
strength of activation of its representation in the child's lexicon
reaches some thoretical threshold that leads to word-form recognition.
This activation may result from the infant being exposed to the actual
word-form, or the result of activation spreading through phonological or
semantic links across lexical representation, as in the case of parallel
activation. The strength of this co-activation is proportional to the
phonological similarity between the two translation equivalents; given
that cognates share higher phonological similarity than non-cognates,
the former should be co-activated more strongly than the latter. This
should lead to a faster accumulation of learning instances for cognates,
compared to non-cognates. Parallel activation would allow bilingual
children to accumulate learning instances for words in both languages
even during fully monolingual situations, but the impact of this
mechanism would be asymmetric across languages: words from the
lower-exposure language would receive stronger activation from words in
the higher-exposure language than vice versa. Therefore, the acquisition
of words from the lower-exposure language would benefit more strongly
from their cognate status than words from the higher-exposure language.
This asymmetric cross-language activation would be consistent with
previous reports of larger priming effects from the dominant to the
non-dominant language.

Consider the example of the Catalan-Spanish cognate translation
equivalent /\textipa{"gat}/--/\textipa{"ga.to}/ {[}\emph{cat}{]}, which
are phonologically very similar. When the child listens to
/\textipa{"gat}/, they will strongly co-activate /\textipa{"ga.to}/ in
parallel. If the child has already formed a form-meaning association for
both word-forms, parallel activation may result from the activation of
their common concept or from activation spreading throught phonological
similarity. We assume semantic co-activation to be constant across
cognate and non-cognate translation equivalents, and focus on
phonological co-activation as an additional source of activation that
affects cognates more strongly than non-cognates. Therefore, this
exposure will count as a learning instance for both co-activated forms.
The case of the non-cognate translation equivalent
/\textipa{"gos-"pe.ro}/ {[}\emph{dog}{]} would be different. Given the
low phonological similarity between both word-forms, an exposure to
\textipa{"gos} will result in a weak activation of /\textipa{"pe.ro}/
leading to such exposure counting as a learning instance for
/\textipa{"gos}/ (which the child was exposed to), but not for
/\textipa{"pe.ro}/. While /\textipa{"gat}/--/\textipa{"ga.to}/ will
benefit from phonological co-activation,
/\textipa{"gos}-\textipa{"pe.ro}/ will not. If the child receives
linguistic input from one of the languages more often than from the
other, this effect might affect each form of the cognate translation
equivalent differently. For instance, if the child receives a larger
amount of Catalan input than Spanish input, they will encounter the
Catalan form /\textipa{"gat}/ more frequently than the Spanish form
/\textipa{"ga.to}/. Through parallel activation, /\textipa{"gat}/ will
activate /\textipa{"ga.to}/ more often than vice versa. Ultimately,
/\textipa{"ga.to}/ will benefit more strongly from its cognate status
than /\textipa{"gat}/, as it receives additional learning instances from
its translation equivalent more often than /\textipa{"gat}/.

To test these predictions, we collected vocabulary data on production
and comprehension from a large sample of bilingual Catalan-Spanish
children. We adopted a Bayesian explanatory item response theory
approach to model the probability of acquisition of 604 Catalan and
Spanish nouns included in the vocabulary checklist. Words were
considered as acquired if caregivers reported such word to be understood
(comprehension) or understood and said (production) by their child. We
estimated the impact of several predictors of interest on the
probability of acquisition, including participants' age and rate of
exposure to the word-form, and the cognate status of the word-form. As
described in the methods section, rate of exposure was a composite
measure taking into account participant' language exposure and word's
lexical frequency. We predicted an interaction between cognate status
and word-form exposure rate in which the probability of comprehension is
higher for low-exposure cognate words, but not for high-exposure cognate
words.

@@

All materials, data, and reproducible code can be found at the OSF
(\url{https://osf.io/hy984/}) and GitHub
(\url{https://github.com/gongcastro/cognate-beginnings}) repositories.
This study was conducted according to guidelines laid down in the
Declaration of Helsinki, and was approved by the Drug Research Ethical
Committee (CEIm) of the IMIM Parc de Salut Mar, reference 2020/9080/I.

@@

To collect vocabulary data from participants, we created an \emph{ad
hoc} questionnaire: the Barcelona Vocabulary Questionnaire (BVQ)
(\protect\hyperlink{ref-garcia-castro2023bvq}{Garcia-Castro et al.,
2023}). This questionnaire was inspired by the MacArthur-Bates
Communicative Development Inventory
(\protect\hyperlink{ref-fenson2007macarthurbates}{Fenson et al., 2007})
and its adaptations to other languages, and was implemented on-line
using the \texttt{formr} platform
(\protect\hyperlink{ref-arslan2020formr}{Arslan et al., 2020}). This
questionnaire is structured in three blocks: (1) a language exposure
questionnaire, (2) a demographic survey, and (3) two vocabulary
checklists. Vocabulary checklists followed a similar structure as the
Oxford Communicative Developmental Inventory
(\protect\hyperlink{ref-hamilton2000infant}{Hamilton et al., 2000}) and
consisted in two lists of words: one in Catalan and one in Spanish. Both
lists included items from a diverse sample of 26 semantic or functional
categories. The Catalan checklist contained 793 items and the Spanish
checklist contained 797. Items in one language were translation
equivalents of the items in the other language (e.g., the same
participant responded to both \emph{gos} and \emph{perro}, Catalan and
Spanish for \emph{dog}), roughly following a one-to-one mapping. Some of
the words in Catalan did not have a clear translation or had more than
one possible translation in Spanish, and vice versa, therefore the
unequal number of words included in the two lists (see
Table~\ref{tbl-items} for a summary of the questionnaire items).

@@

\centering
\begin{tabular}{lccccl}
\toprule
Semantic category & List A & List B & List C & List D & Examples\\
\midrule
Household items & 31 & 26 & 30 & 25 & clock, video\\
Food and drink & 29 & 26 & 23 & 27 & sausage, yogurt\\
Animals & 26 & 23 & 19 & 25 & panther, tiger\\
Outside & 14 & 13 & 13 & 15 & farm, stone\\
Body parts & 14 & 12 & 11 & 11 & face, finger\\
\addlinespace
Toys & 11 & 11 & 12 & 13 & piano, racket\\
Clothes & 12 & 12 & 10 & 10 & zipper, sandal\\
Vehicles & 9 & 10 & 11 & 10 & helicopter, tractor\\
Colours & 6 & 6 & 6 & 6 & red, green\\
People & 7 & 4 & 6 & 6 & police, babysitter\\
\addlinespace
Furniture and rooms & 4 & 4 & 4 & 4 & corridor, terrace\\
Time & 2 & 2 & 2 & 2 & day, night\\
Adventures & 1 & 1 & 1 & 1 & witch\\
Parts of things & 1 & 1 & 1 & 1 & wheel\\
\midrule
N & 167 & 151 & 149 & 156 & -\\
\bottomrule
\end{tabular}
\end{table}

For each word included in the vocabulary checklists, we asked parents to
report whether their child was able to understand it, understand
\emph{and} say it, or did not understand or say it (checked out by
default). Given the large number of words in the vocabulary checklists,
we created four different subsets of the complete list of items: A, B,
C, and D. Each subset contained a random but representative sub-sample
of the items from the complete list (see Table~\ref{tbl-items}).
Semantic or functional categories with less than 16 items---thus
resulting in less than four items after dividing it in four lists---were
not divided: all of their items were included in the four lists. Items
that were part of the trial lists of some ongoing experiments in the lab
were also included in all versions. The resulting reduced list contained
between 343 and 349 Catalan words, and between 349 and 371 Spanish
words. Participants were randomly allocated into one of the four
subsets.

@@

We collected 440 responses to the questionnaire from 369 distinct
children from the Metropolitan Area of Barcelona between Marth 30th,
2020 and October 31th, 2022: 314 of those participants participated
once, 43 twice, 8 three times, and 4 four times. Recurrent participants
provided responses with a minimum of 25 days between responses, and a
maximum of 527 days, and were always allocated to the same questionnaire
list (A, B, C, or D). Participants were part of the database of the
Laboratori de Recerca en Infància (Universitat Pompeu Fabra) and were
contacted by e-mail or phone if their child was aged between 12 and 32
months, and had not been reported to be exposed more than 10\% of the
time to a language other than Spanish or Catalan (see
Table~\ref{tbl-participants} for a more detailed description of the
sample). In total, 70 participants (15.91\%) participants were reported
to be exposed in less than 10\% to a third language other than Catalan
and Spanish. All families provided informed consent before
participating. Upon consent, families were sent a link to the
questionnaire via e-mail, which they filled from a computer, laptop, or
mobile device. Filling the questionnaire took 30 minutes approximately.
After completion, families were rewarded with a token of appreciation.

@@

We implemented the model using \texttt{brms}
(\protect\hyperlink{ref-burkner2017brms}{Bürkner, 2017}), a R interface
to the Stan probabilistic language (2.32.1)
(\protect\hyperlink{ref-carpenter2017stan}{Carpenter et al., 2017}). We
ran four iteration chains using the by-default No U-Turn Sampler
algorithm with 1,000 iterations each and an additional 1,000 warm-up
iterations per chain.

@@

>>>>

supported by the European Union's Horizon2020 research and innovation
program under Marie Skłodowska--Curie Grant (765556) and partially
supported by Portuguese national funds through the Foundation for
Science and Technology, under project UIDB/00214/2020, awarded to the
Center of Linguistics of the University of Lisbon. IC was supported by
the Investigo program funded by the European Union's NextGenerationEU
(NGEU) recovery plan. NSG was supported by an ICREA Academia 2019 award
from the Catalan Institution for Research and Advanced Studies (ICREA).
We are grateful to Chiara Santolin, Ege E. Özer, and the rest of the
Speech Acquisition and Perception research group, and to Alicia
Franco-Martínez and Cristina Rodríguez-Prada, for their helpful
feedback. We thank Xavier Mayoral, Silvia Blanch, and Cristina Cuadrado
for their technical support, and Cristina Dominguez and Katia Pistrin
for their efforts in recruiting infants. We also thank all families and
infants who participated in the experiments.}}

@@

08005\\\texttt{\href{mailto:daniela.avila@upf.edu}{daniela.avila@upf.edu}}\And

@@

questionnaire (BVQ), to collect vocabulary data from 369 Catalan-Spanish

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, frame hidden, sharp corners, interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi

@@

Floccia et al. (\protect\hyperlink{ref-floccia2018introduction}{2018})
published an impressive study comparing vocabularies of children
learning several language pairs differing in their percentage of
cognates. The authors collected vocabulary data on word comprehension
and production from 372 24-month-old bilingual toddlers living in the
United Kingdom who were learning English and an additional language. The
additional language was one of 13 typologically diverse languages:
Bengali, Cantonese Chinese, Dutch, French, German, Greek, Hindi or Urdu,
Italian, Mandarin Chinese, Polish, Portuguese, Spanish, and Welsh. The
authors calculated the average lexical overlap between the words in each
of these additional languages and their translation equivalents in
English. Lexical overlap was calculated in terms of phonological
similarity (described below) and it was taken as a proxy of the degree
of cognateness between each pair of languages. Floccia and co-workers
reported an increase in vocabulary size in the additional language
(i.e., not English) associated with an increase in the average
phonological similarity between the translation equivalents of each
language pair. For example, English-Dutch bilinguals (languages with a
high phonological overlap), were able to produce more Dutch words than
English-Mandarin bilinguals (languages with a low phonological overlap)
were able to produce in Mandarin. Blom et al.
(\protect\hyperlink{ref-blom2020crosslanguage}{2020}), Bosma et al.
(\protect\hyperlink{ref-bosma2019longitudinal}{2019}), and Gampe et al.
(\protect\hyperlink{ref-gampe2021does}{2021}) reported similar results,
providing converging evidence of a facilitatory effect of a lower
language distance (i.e., higher degree of cognateness) on vocabulary
size.

A second set of studies suggested that cognates are overrepresented in
bilinguals' early lexicon. Bosch \& Ramon-Casas
(\protect\hyperlink{ref-bosch2014first}{2014}) collected parental
reports of expressive vocabulary from 48 Catalan-Spanish bilinguals aged
18 months and found that cognates represented a larger proportion of
vocabulary than non-cognates. Schelletter
(\protect\hyperlink{ref-schelletter2002effect}{2002}) provided
converging evidence from a longitudinal single-case study, in which an
English-German bilingual child produced cognates earlier than
non-cognates, on average. But the high proportion of cognates in the
vocabulary of the participants in these two studies may not necessarily
evidence of a facilitation effect of cognateness, but rather of simply
the high proportion of cognates present in the pair of languages being
learned. For instance, if two given languages share a high proportion of
cognates like 70\%, the vocabulary contents of children learning both
languages should, in principle, approximate such proportion of cognates,
even in the absence of a cognateness facilitation effect. More recently,
Mitchell et al. (\protect\hyperlink{ref-mitchell2022cognates}{2022})
addressed this issue in a longitudinal study. The authors collected
expressive vocabulary data of 47 16- to 30-month-old French-English
bilinguals living in Canada, in both languages. They created two lists
of translation equivalents; one made of 131 cognates, and one made of
406 non-cognates. Across ages, the proportion of words that children
were reportedly able to produce was higher in the cognate lists than in
the non-cognate list. Critically, this difference persisted after both
lists were matched in size, controlling their semantic category (i.e.,
furniture, animals, food were similarity represented in both lists) and
age-of-acquisition norms (an index of word difficulty). Taken together,
the results of these two lines of research support the hypothesis that
phonological similarity (as reflected in cognateness) plays a
facilitation role in bilingual word learning.

Parallel activation of bilinguals' lexicons has been proposed as the
underlying mechanism for such facilitatory effect (e.g.,
\protect\hyperlink{ref-floccia2018introduction}{Floccia et al., 2018};
\protect\hyperlink{ref-mitchell2022cognates}{Mitchell et al., 2022}).
The parallel activation hypothesis stems from the language non-selective
account of lexical access, which suggests that bilinguals activate both
languages simultaneously during language processing, even in fully
monolingual contexts. Evidence with adult bilinguals supporting the
language-non selective account of lexical access has been reported for
language comprehension and production, across the auditory and visual
(reading and signing) modalities
(\protect\hyperlink{ref-gimeno-martinez2021crosslinguistic}{Gimeno-Martínez
et al., 2021}; \protect\hyperlink{ref-hoshino2008cognate}{Hoshino \&
Kroll, 2008}; \protect\hyperlink{ref-morford2011when}{Morford et al.,
2011}; \protect\hyperlink{ref-shook2012bimodal}{Shook \& Marian, 2012};
\protect\hyperlink{ref-spivey1999cross}{Spivey \& Marian, 1999}; see
\protect\hyperlink{ref-kroll2017bilingual}{Kroll \& Ma, 2017} for
review). One of the clearest pieces of evidence of parallel activation
was provided by Costa et al.
(\protect\hyperlink{ref-costa2000cognate}{2000}). In this study, Spanish
monolinguals and Catalan-Spanish bilingual adults were asked to name
pictures of common objects in Spanish. In half of the trials, the object
labels were cognates in Spanish and Catalan (\emph{árbol}-\emph{arbre},
translations of \emph{tree}), whereas in the other half of the trial
labels were non-cognates (\emph{mesa}-\emph{taula}, translations of
\emph{table}). Obviously, such distinction was only relevant for
bilinguals. Bilinguals named cognate pictures faster than non-cognate
pictures, even after adjusting for the lexical frequency of the items.
In contrast, Spanish monolinguals, who were unfamiliar with the Catalan
translations of the Spanish words they uttered, showed equivalent naming
times for the two types of stimuli. The authors interpreted the
difference between cognates and non-cognates in bilinguals as reflecting
the additional phonological activation that cognate words would receive
from their translation equivalents (due to language non-selective
activation of bilinguals' lexicons). These results showed that
bilinguals' Catalan phonology was activated during the production of
Spanish words, facilitating the naming of cognate pictures. More
recently, evidence of parallel activation has been reported in bilingual
toddlers and children too
(\protect\hyperlink{ref-bosma2020cognate}{Bosma \& Nota, 2020};
\protect\hyperlink{ref-floccia2020translation}{Floccia et al., 2020};
\protect\hyperlink{ref-poarch2012crosslanguage}{Poarch \& Hell, 2012};
\protect\hyperlink{ref-poulin-dubois2013lexical}{Poulin-Dubois et al.,
2013}; \protect\hyperlink{ref-vonholzen2019impact}{Von Holzen et al.,
2019}; \protect\hyperlink{ref-vonholzen2012language}{Von Holzen \& Mani,
2012}). Although there is a consensus on the role of parallel activation
in bilinguals' lexical processing and acquisition, previous studies do
not address its influence on the learning trajectories of individual
words. Results are aggregated across words and provide no information
about the specific dynamics of how parallel activation influences word
learning. This is the goal of the present research.

We propose an account in which a learning instance for a word may also
represent a learning instance for its translation equivalent, to the
extent that such translation equivalent is co-activated. We use the term
\emph{learning instance} in the fashion of accumulator models of
language acquisition; as an exposure to a word-form that constitutes an
opportunity for the child to accumulate information about the word. We
do not assume if a learning instance is a discrete or a continuous unit
of accumulation of information. We consider that a learning instance of
a word is an exposure to its (phonological) form if the resulting
strength of activation of its representation in the child's lexicon
reaches some thoretical threshold that leads to word-form recognition.
This activation may result from the infant being exposed to the actual
word-form, or the result of activation spreading through phonological or
semantic links across lexical representation, as in the case of parallel
activation. The strength of this co-activation is proportional to the
phonological similarity between the two translation equivalents; given
that cognates share higher phonological similarity than non-cognates,
the former should be co-activated more strongly than the latter. This
should lead to a faster accumulation of learning instances for cognates,
compared to non-cognates. Parallel activation would allow bilingual
children to accumulate learning instances for words in both languages
even during fully monolingual situations, but the impact of this
mechanism would be asymmetric across languages: words from the
lower-exposure language would receive stronger activation from words in
the higher-exposure language than vice versa. Therefore, the acquisition
of words from the lower-exposure language would benefit more strongly
from their cognate status than the acquisition of words from the
higher-exposure language. This asymmetric cross-language activation
would be consistent with previous reports of larger priming effects from
the dominant to the non-dominant language (e.g.,
\protect\hyperlink{ref-grainger1998masked}{Grainger, 1998}).

Consider the example of the Catalan-Spanish cognate translation
equivalent /\textipa{"gat}/--/\textipa{"ga.to}/ {[}\emph{cat}{]}, which
are phonologically very similar. When the child listens to
/\textipa{"gat}/, they will strongly co-activate /\textipa{"ga.to}/ in
parallel. If the child has already formed a form-meaning association for
both word-forms, parallel activation may result from the activation of
their common concept or from activation spreading throught phonological
similarity. We assume semantic co-activation to be constant across
cognate and non-cognate translation equivalents, and focus on
phonological co-activation as an additional source of activation that
affects cognates more strongly than non-cognates. Therefore, this
exposure will count as a learning instance for both co-activated forms.
The case of the non-cognate translation equivalent
/\textipa{"gos}/--/\textipa{"pe.ro}/ {[}\emph{dog}{]} would be
different. Given the low phonological similarity between both
word-forms, an exposure to \textipa{"gos} will result in a weak
activation of /\textipa{"pe.ro}/ leading to such exposure counting as a
learning instance for /\textipa{"gos}/ (which the child was exposed to),
but not for /\textipa{"pe.ro}/. While
/\textipa{"gat}/--/\textipa{"ga.to}/ will benefit from phonological
co-activation, /\textipa{"gos}--\textipa{"pe.ro}/ will not. If the child
receives linguistic input from one of the languages more often than from
the other, this effect might affect each form of the cognate translation
equivalent differently. For instance, if the child receives a larger
amount of Catalan input than Spanish input, they will encounter the
Catalan form /\textipa{"gat}/ more frequently than the Spanish form
/\textipa{"ga.to}/. Through parallel activation, /\textipa{"gat}/ will
activate /\textipa{"ga.to}/ more often than vice versa. Ultimately,
/\textipa{"ga.to}/ will benefit more strongly from its cognate status
than /\textipa{"gat}/, as it receives additional learning instances from
its translation equivalent more often than /\textipa{"gat}/.

To test these predictions, we collected vocabulary data on production
and comprehension from a large sample of bilingual Catalan-Spanish
children. We adopted a Bayesian explanatory item response theory
approach to model the probability of acquisition of 604 Catalan and
Spanish nouns included in the vocabulary checklist. Words were
considered as acquired if caregivers reported such word to be understood
(comprehension) or understood and said (production) by their child. We
estimated the impact of several predictors of interest on the
probability of acquisition, including participants' age and rate of
exposure to the word-form, and the cognate status of the word-form. As
described in the Methods section, rate of exposure was a composite
measure taking into account participant' language exposure and word's
lexical frequency. We predicted an interaction between cognate status
and word-form exposure rate in which the probability of comprehension is
higher for low-exposure cognate words, but not for high-exposure cognate
words.

@@

All materials, data, and reproducible code can be found at the OSF
(\url{https://osf.io/hy984/}) and GitHub
(\url{https://github.com/gongcastro/cognate-beginnings}) repositories.
For reproducibility, a Docker image of the RStudio session is available
on
DockerHub(\url{https://hub.docker.com/repository/docker/gongcastro/cognate-beginnings/}).
This study was conducted according to guidelines laid down in the
Declaration of Helsinki, and was approved by the Drug Research Ethical
Committee (CEIm) of the IMIM Parc de Salut Mar, reference 2020/9080/I.

@@

To collect vocabulary data from participants, we created an \emph{ad
hoc} questionnaire: the Barcelona Vocabulary Questionnaire (BVQ)
(\protect\hyperlink{ref-garcia-castro2023bvq}{Garcia-Castro et al.,
2023}). This questionnaire was inspired by the MacArthur-Bates
Communicative Development Inventory
(\protect\hyperlink{ref-fenson2007macarthurbates}{Fenson et al., 2007})
and its adaptations to other languages, and was implemented on-line
using the \texttt{formr} platform
(\protect\hyperlink{ref-arslan2020formr}{Arslan et al., 2020}). This
questionnaire is structured in three blocks: (1) a language exposure
questionnaire, (2) a demographic survey, and (3) two vocabulary
checklists. Vocabulary checklists followed a similar structure as the
Oxford Communicative Developmental Inventory
(\protect\hyperlink{ref-hamilton2000infant}{Hamilton et al., 2000}) and
consisted of two lists of words, one in Catalan and one in Spanish. Both
lists included items from a diverse sample of 26 semantic or functional
categories. The Catalan checklist contained 793 items and the Spanish
checklist contained 797. Items in one language were translation
equivalents of the items in the other language (e.g., the same
participant responded to both \emph{gos} and \emph{perro}, Catalan and
Spanish for \emph{dog}), roughly following a one-to-one mapping. Some of
the words in Catalan did not have a clear translation or had more than
one possible translation in Spanish, and vice versa, therefore the
unequal number of words included in the two lists.

@@

\centering
\begin{tabular}{lccccl}
\toprule
Semantic category & List A & List B & List C & List D & Examples\\
\midrule
Household items & 31 & 26 & 30 & 25 & clock, video\\
Food and drink & 29 & 26 & 23 & 27 & sausage, yogurt\\
Animals & 26 & 23 & 19 & 25 & panther, tiger\\
Outside & 14 & 13 & 13 & 15 & farm, stone\\
Body parts & 14 & 12 & 11 & 11 & face, finger\\
\addlinespace
Toys & 11 & 11 & 12 & 13 & piano, racket\\
Clothes & 12 & 12 & 10 & 10 & zipper, sandal\\
Vehicles & 9 & 10 & 11 & 10 & helicopter, tractor\\
Colours & 6 & 6 & 6 & 6 & red, green\\
People & 7 & 4 & 6 & 6 & police, babysitter\\
\addlinespace
Furniture and rooms & 4 & 4 & 4 & 4 & corridor, terrace\\
Time & 2 & 2 & 2 & 2 & day, night\\
Adventures & 1 & 1 & 1 & 1 & witch\\
Parts of things & 1 & 1 & 1 & 1 & wheel\\
\midrule
*N* & 167 & 151 & 149 & 156 & -\\
\bottomrule
\end{tabular}
\end{table}

For each word included in the vocabulary checklists, we asked parents to
report whether their child was able to understand it, understand
\emph{and} say it, or did not understand or say it (checked out by
default). Given the large number of words in the vocabulary checklists,
we created four different subsets of the complete list of items (A, B,
C, and D) Each subset contained a random but representative sub-sample
of the items from the complete list (see Table~\ref{tbl-items}).
Semantic or functional categories with less than 16 items---thus
resulting in less than four items after dividing it in four
subsets---were not divided: all of their items were included in the four
subsets. Items that were part of the trial lists of some ongoing
experiments in the lab were also included in all versions. The resulting
reduced list contained between 343 and 349 Catalan words, and between
349 and 371 Spanish words.

@@

We collected 436 questionnaire responses from 366 distinct children (175
female, 179 male, 12 not reported, mainly White). Participants were aged
12-32 months (\emph{M} = 22.23, \emph{SD} = 4.88, \emph{Range} =
12.06-31.93). Of those participants, four participated four times, eight
participated three times, 42 participated twice, and 312 participated
once. Recurrent participants provided responses with a minimum of 25
days between responses, and a maximum of 527 days. Participants were
randomly allocated into one of the four questionnaire subsets (A, B, C,
or D). Each participant was always allocated to the same subset.

Participants were residents in the Metropolitan Area of Barcelona
(Catalonia, Spain). Data collection took place between March 30th, 2020
and October 31th, 2022. Participants were part of the database of the
Laboratori de Recerca en Infància at the Universitat Pompeu Fabra, and
were contacted by e-mail or phone if their child was aged between 12 and
32 months, and had not been reported to be exposed more than 10\% of the
time to a language other than Spanish or Catalan (see
Table~\ref{tbl-participants} for a more detailed description of the
sample). In total, 70 participants (16.06\%) participants were reported
to be exposed in less than 10\% to a third language other than Catalan
and Spanish. All families provided informed consent before
participating. Upon consent, families were sent a link to the
questionnaire via e-mail, which they filled from a computer, laptop, or
mobile device. Filling the questionnaire took 30 minutes approximately.
After completion, families were rewarded with a token of appreciation.

@@

We implemented the model using \texttt{brms}
(\protect\hyperlink{ref-burkner2017brms}{Bürkner, 2017}), an R interface
to the Stan probabilistic language (2.32.1)
(\protect\hyperlink{ref-carpenter2017stan}{Carpenter et al., 2017}). We
ran four iteration chains using the by-default No U-Turn Sampler
algorithm with 1,000 iterations each and an additional 1,000 warm-up
iterations per chain.

@@

We now present our confirmatory analysis, and report a summary of the
regression coefficients of interest. Table~\ref{tbl-coefs} shows the
summary of the posterior distribution of the fixed regression
coefficients, and their degree of overlap with the ROPE. For
interpretability, we report the estimated regression coefficients
transformed to the probability scale. The resulting values correspond to
the maximum difference in probability of acquisition
(\emph{Comprehension} or \emph{Comprehension and Production}) equivalent
to a one standard deviation change in each predictor.
